\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-18 k-planar.tex
added Brandenburg etc references and sentences

This result is used below to obtain product structure theorems for 1-planar multigraphs and for map graphs.
% \todo{Move this comment to the section on $1$-planar graphs?}
For simple 1-planar graphs it is essential that we allow frames with parallel edges.
% \note{DW}{I feel that at this point we need to justify the introduction of multigraphs.}

@@

\cref{1-planar} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma, which is of independent interest.

\note{DW}{Say something about how \cref{1_planar_is_4_framed} is better than similar results in the literature.}

\note{PM}{See my attempt below.}
% \note{PM}{I don't know what to say without explaining that $4$-framed (simple) graphs only contain $3$-connected $1$-planar graphs.  Vida's example shows that there is a $1$-plane embedding of a graph that can't be augmented into a $4$-framed graph, but I haven't checked if it can't be drawn some better way. And I also don't know if the Bekos claim that $8$-framed (smple) graphs contain all $1$-planar graphs is proven anywhere and, if so, is it tight?}

@@

Before proving \cref{1_planar_is_4_framed}, we remark that the lemma seems not to be true if we replace `multigraph' with `graph'.  We are only aware of two previous results on the relationship between $1$-planar graphs and $d$-framed graphs \cite{BDGGMR}:
\begin{inparaenum}[(i)]
  \item Every $3$-connected $1$-planar graph is contained in some $4$-framed graph.
  \item Every $1$-planar graph is contained in some $8$-framed graph.
\end{inparaenum}
Neither of these results allows a proof of \cref{1-planar} without sacrificing generality or accepting a larger constant.

\begin{proof}[Proof of \cref{1_planar_is_4_framed}]
Let $G$ be a $1$-plane graph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}.
  % We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.
We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

@@

Prior to the current work, the strongest known upper bounds on the $p$-centered chromatic number of $(g,k)$-planar graphs $G$ were doubly-exponential in $p$, as we now explain. \citet{dujmovic.eppstein.ea:structure} proved that $G$ has layered treewidth $(4g+6)(k+1)$.
Van den Heuvel and Wood~\citep{vdHW17} showed that this implies that $G$  has $r$-strong colouring number at most $(4g + 6)(k + 1)(2r + 1)$. By a result of \citet{zhu:colouring}, $G$ has $r$-weak colouring number at most $( (4g + 6)(k + 1)(2r + 1) )^r$, which by another result of  \citet{zhu:colouring} implies that $G$ has  $p$-centered chromatic number at most $( (4g+6)(k+1)(2^{p-1} + 1) )^{2^{p-2}}$. The above results are substantial improvements, providing bounds on $\chi_p(G)$ that are polynomial in $p$ for fixed $g$ and $k$.

@@

\note{DW}{This subsection needs a clean-up; define $(g,d)$-framed graph. Is the following lemma known (at least in the $g=0$ case)?}
\note{PM}{I have no idea if this is known.  If so, it's probably not stated in terms of frames.}
\begin{lem}
\label{MewMapGraphLemma}
For all integers $g\geq 0$ and $d\geq 3$, every $(g,d)$-map graph is a spanning subgraph of a $(g,d)$-framed graph.
\end{lem}

@@

\begin{thm}
\label{NewMapGraph}
Every $d$-map graph is contained in $H \boxtimes P \boxtimes K_{ d + 3\floor{d/2} -3 }$ for some graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

Map graphs are defined as follows. Start with a graph $G_0$ embedded in a surface of Euler genus $g$, with each face labelled a `nation' or a `lake', where each vertex of $G_0$ is incident with at most $d$ nations. Let $G$ be the graph whose vertices are the nations of $G_0$, where two vertices are adjacent in $G$ if the corresponding faces in $G_0$ share a vertex. Then $G$ is called a \defin{$(g,d)$-map graph}.  A $(0,d)$-map graph is called a (plane) \defin{$d$-map graph}; see \citep{FLS-SODA12,CGP02} for example.
The $(g,3)$-map graphs are precisely the graphs of Euler genus at most $g$; see \citep{dujmovic.eppstein.ea:structure}.
%So $(g,d)$-map graphs generalise graphs embedded in a surface, and we now assume that $d\geq 4$ for the remainder of this section.

@@

% \note{DW}{\cref{MapPartition} has been updated.}

>>>>

This result is used below to obtain product structure theorems for 1-planar multigraphs and for map graphs. 
For simple 1-planar graphs it is essential that we allow frames with parallel edges.

@@

\cref{1-planar} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma. Similar results appear in earlier works \cite{CGP06,BDGGMR,Brandenburg19,Brandenburg20} but we state the precise lemma and provide a proof here for the sake of completeness.

@@

\note{DW}{Write  ``A multigraph $G$ is $1$-planar if and only if ...''}

%Before proving \cref{1_planar_is_4_framed}, we remark that the lemma seems not to be true if we replace `multigraph' with `graph'.  We are only aware of two previous results on the relationship between $1$-planar graphs and $d$-framed graphs \cite{BDGGMR}:
%\begin{inparaenum}[(i)]
%  \item Every $3$-connected $1$-planar graph is contained in some $4$-framed graph.
%  \item Every $1$-planar graph is contained in some $8$-framed graph.
%\end{inparaenum}
%Neither of these results allows a proof of \cref{1-planar} without sacrificing generality or accepting a larger constant.

\begin{proof}
Let $G$ be a $1$-plane graph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}.
  % We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.
We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

@@

Prior to the current work, the strongest known upper bounds on the $p$-centered chromatic number of $(g,k)$-planar graphs $G$ were doubly-exponential in $p$, as we now explain. \citet{dujmovic.eppstein.ea:structure} proved that $G$ has layered treewidth $(4g+6)(k+1)$.
Van den Heuvel and Wood~\citep{vdHW18} showed that this implies that $G$  has $r$-strong colouring number at most $(4g + 6)(k + 1)(2r + 1)$. By a result of \citet{zhu:colouring}, $G$ has $r$-weak colouring number at most $( (4g + 6)(k + 1)(2r + 1) )^r$, which by another result of  \citet{zhu:colouring} implies that $G$ has  $p$-centered chromatic number at most $( (4g+6)(k+1)(2^{p-1} + 1) )^{2^{p-2}}$. The above results are substantial improvements, providing bounds on $\chi_p(G)$ that are polynomial in $p$ for fixed $g$ and $k$.

@@

Map graphs are defined as follows. Start with a graph $G_0$ embedded in a surface of Euler genus $g$, with each face labelled a `nation' or a `lake', where each vertex of $G_0$ is incident with at most $d$ nations. Let $G$ be the graph whose vertices are the nations of $G_0$, where two vertices are adjacent in $G$ if the corresponding faces in $G_0$ share a vertex. Then $G$ is called a \defin{$(g,d)$-map graph}.  A $(0,d)$-map graph is called a (plane) \defin{$d$-map graph}; see \citep{FLS-SODA12,CGP02} for example. The $(g,3)$-map graphs are precisely the graphs of Euler genus at most $g$; see \citep{dujmovic.eppstein.ea:structure}.

\note{DW}{This subsection needs a clean-up; define $(g,d)$-framed graph.}

\begin{thm}
\label{NewMapGraph}
Every $d$-map graph is contained in $H \boxtimes P \boxtimes K_{ d + 3\floor{d/2} -3 }$ for some planar graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

\cref{NewMapGraph} is an immediate consequence of \cref{d_framed_product_stucture}} and the next lemma. Similar results to \cref{MewMapGraphLemma} appear in earlier works \cite{CGP06,BDGGMR,Brandenburg19,Brandenburg20} but we state the precise lemma and provide a proof here for the sake of completeness.

\begin{lem}
\label{MewMapGraphLemma}
For all integers $g\geq 0$ and $d\geq 3$, every $(g,d)$-map graph is a spanning subgraph of a $(g,d)$-framed graph.
\end{lem}

@@

\note{DW}{give a better version of \cref{MapPartition} with $g=0$. }

%So $(g,d)$-map graphs generalise graphs embedded in a surface, and we now assume that $d\geq 4$ for the remainder of this section.

@@
